---
title: "Project 8 Template"
output: pdf_document
---

```{r}
# Add to this package list for additional SL algorithms
pacman::p_load(
  tidyverse,
  ggthemes,
  ltmle,
  tmle,
  SuperLearner,
  tidymodels,
  caret,
  dagitty,
  ggdag,
  here)

heart_disease <- read_csv(here('heart_disease_tmle.csv'))
```

# Introduction

Heart disease is the leading cause of death in the United States, and treating it properly is an important public health goal. However, it is a complex disease with several different risk factors and potential treatments. Physicians typically recommend changes in diet, increased exercise, and/or medication to treat symptoms, but it is difficult to determine how effective any one of these factors is in treating the disease. In this project, you will explore SuperLearner, Targeted Maximum Likelihood Estimation (TMLE), and Longitudinal Targeted Maximum Likelihood Estimation (LTMLE). Using a simulated dataset, you will explore whether taking blood pressure medication reduces mortality risk. 

# Data

This dataset was simulated using R (so it does not come from a previous study or other data source). It contains several variables:

\begin{itemize}
    \item \textbf{blood\_pressure\_medication}: Treatment indicator for whether the individual took blood pressure medication (0 for control, 1 for treatment)
    \item \textbf{mortality}: Outcome indicator for whether the individual passed away from complications of heart disease (0 for no, 1 for yes)
    \item \textbf{age}: Age at time 1
    \item \textbf{sex\_at\_birth}: Sex assigned at birth (0 female, 1 male)
    \item \textbf{simplified\_race}: Simplified racial category. (1: White/Caucasian, 2: Black/African American, 3: Latinx, 4: Asian American, \newline 5: Mixed Race/Other)
    \item \textbf{income\_thousands}: Household income in thousands of dollars
    \item \textbf{college\_educ}: Indicator for college education (0 for no, 1 for yes)
    \item \textbf{bmi}: Body mass index (BMI)
    \item \textbf{chol}: Cholesterol level
    \item \textbf{blood\_pressure}: Systolic blood pressure 
    \item \textbf{bmi\_2}: BMI measured at time 2
    \item \textbf{chol\_2}: Cholesterol measured at time 2
    \item \textbf{blood\_pressure\_2}: BP measured at time 2
    \item \textbf{blood\_pressure\_medication\_2}: Whether the person took treatment at time period 2 
\end{itemize}

For the "SuperLearner" and "TMLE" portions, you can ignore any variable that ends in "\_2", we will reintroduce these for LTMLE.

# SuperLearner

## Modeling

Fit a SuperLearner model to estimate the probability of someone dying from complications of heart disease, conditional on treatment and the relevant covariates. Do the following:

\begin{enumerate}
    \item Choose a library of at least 5 machine learning algorithms to evaluate. \textbf{Note}: We did not cover how to hyperparameter tune constituent algorithms within SuperLearner in lab, but you are free to do so if you like (though not required to for this exercise). 
    \item Split your data into train and test sets.
    \item Train SuperLearner
    \item Report the risk and coefficient associated with each model, and the performance of the discrete winner and SuperLearner ensemble
    \item Create a confusion matrix and report your overall accuracy, recall, and precision
\end{enumerate}

```{r}
# Fit SuperLearner Model

listWrappers() # check the available algorithms

## sl lib

sl_libs <- c('SL.glmnet', 'SL.glm', 'SL.lda','SL.ranger','SL.svm')

## Train/Test split

# initial_split function from tidymodels/rsample
hd_split <- initial_split(heart_disease, prop = 3/4)

# Declare the training set with rsample::training()
train <- training(hd_split)

y_train <- train %>% 
  pull(mortality) 

# x_train is everything but the outcome  
x_train <- train %>%
  select(-mortality)

# Do the same procedure with the test set
test <- testing(hd_split)

y_test <- test %>%
  pull(mortality)

x_test <- test %>%
  select(-mortality)

## Train SuperLearner

sl = SuperLearner(Y = y_train,
                  X = x_train,
                  family = binomial(),
                  SL.library = sl_libs)

## Risk and Coefficient of each model

sl

## Discrete winner and superlearner ensemble performance

sl$cvRisk[which.min(sl$cvRisk)]

## Confusion Matrix

preds <- predict(sl,
                 x_test,
                 onlySL = TRUE)

# start with y_test
validation <- y_test %>%
  # add our predictions
  bind_cols(preds$pred[,1]) %>%
  # rename columns
  rename(obs = `...1`,
         pred = `...2`) %>%
  mutate(pred = ifelse(pred >= .5, 
                           1,
                           0))

head(validation)

caret::confusionMatrix(as.factor(validation$pred),
                       as.factor(validation$obs))

```

## Discussion Questions

\begin{enumerate}
    \item Why should we, in general, prefer the SuperLearner ensemble to the discrete winner in cross-validation? Or in other words, what is the advantage of "blending" algorithms together and giving them each weights, rather than just using the single best algorithm (with best being defined as minimizing risk)?
    \item We prefer the Superlearner ensemble to the discrete winner because the former produces an algorithm that is a weighted combination of machine learning algorithms, which has been shown to perform at least as well as any of the constituent algorithms. This blended algorithm may be more optimal for a particular dataset than a single algorithm, essentially because it is data-adaptive (tailored to the particular dataset) while still maintaining the impressive out-of-sample performance that makes machine learning so popular.
\end{enumerate}


# Targeted Maximum Likelihood Estimation

## Causal Diagram

TMLE requires estimating two models:

\begin{enumerate}
    \item The outcome model, or the relationship between the outcome and the treatment/predictors, $P(Y|(A,W)$.
    \item The propensity score model, or the relationship between assignment to treatment and predictors $P(A|W)$
\end{enumerate}

Using ggdag and daggity, draw a directed acylcic graph (DAG) that describes the relationships between the outcome, treatment, and covariates/predictors. Note, if you think there are covariates that are not related to other variables in the dataset, note this by either including them as freestanding nodes or by omitting them and noting omissions in your discussion.

```{r}
# DAG for TMLE
dag<-dagitty('dag {
bb="0,0,1,1"
"BMI T1" [pos="0.628,0.518"]
"BP Medication T1" [exposure,pos="0.162,0.730"]
"Blood Pressure T1" [pos="0.413,0.588"]
"Cholesterol T1" [pos="0.764,0.429"]
"Race/Ethnicity" [pos="0.219,0.224"]
"Sex at birth" [pos="0.669,0.205"]
Age [pos="0.118,0.337"]
Education [pos="0.462,0.288"]
Income [pos="0.167,0.505"]
Mortality [outcome,pos="0.674,0.737"]
"BMI T1" -> "Blood Pressure T1"
"BMI T1" -> "Cholesterol T1"
"BMI T1" -> Mortality
"BP Medication T1" -> Mortality
"Blood Pressure T1" -> "BP Medication T1"
"Blood Pressure T1" -> Mortality
"Cholesterol T1" -> "Blood Pressure T1"
"Cholesterol T1" -> Mortality
"Race/Ethnicity" -> "BMI T1"
"Race/Ethnicity" -> "BP Medication T1"
"Race/Ethnicity" -> "Blood Pressure T1"
"Race/Ethnicity" -> "Cholesterol T1"
"Race/Ethnicity" -> Education
"Race/Ethnicity" -> Income
"Race/Ethnicity" -> Mortality
"Sex at birth" -> "BMI T1"
"Sex at birth" -> "BP Medication T1"
"Sex at birth" -> "Blood Pressure T1"
"Sex at birth" -> "Cholesterol T1"
"Sex at birth" -> Education
"Sex at birth" -> Income
"Sex at birth" -> Mortality
Age -> "Blood Pressure T1"
Age -> "Cholesterol T1"
Age -> Education
Age -> Income
Age -> Mortality
Education -> "BMI T1"
Education -> "BP Medication T1"
Education -> "Blood Pressure T1"
Education -> "Cholesterol T1"
Education -> Income
Education -> Mortality
Income -> "BMI T1"
Income -> "BP Medication T1"
Income -> "Blood Pressure T1"
Income -> "Cholesterol T1"
Income -> Mortality
}')
plot(dag)


```

### Discussion 

Accodring to our DAG, the minimal sufficient adjustment set of covariates for evaluating the relationship between Blood Pressure Medication and Mortality are; *Blood pressure, Education, Income, Simplified Race/Ethnicity, and Sex at birth*. When we control for these five covariates in our model, we are able to ensure that all backdoor paths are blocked and we are able to fulfill the independence assumption. 

## TMLE Estimation

Use the `tmle` package to estimate a model for the effect of blood pressure medication on the probability of mortality. Do the following:

\begin{enumerate}
    \item Use the same SuperLearner library you defined earlier
    \item Use the same outcome model and propensity score model that you specified in the DAG above. If in your DAG you concluded that it is not possible to make a causal inference from this dataset, specify a simpler model and note your assumptions for this step.
    \item Report the average treatment effect and any other relevant statistics
\end{enumerate}

```{r}

# new SL library
sl_libs <- c('SL.glmnet', 'SL.glm','SL.ranger', 'SL.step', 'SL.rpartPrune')

# specify covariates needed for analyses
W<-subset(heart_disease, select=c(sex_at_birth,simplified_race,income_thousands,college_educ, blood_pressure))

TMLE.est = tmle(Y = heart_disease$mortality, 
                A = heart_disease$blood_pressure_medication,
                W = W,
                Q.SL.library = sl_libs,
                g.SL.library = sl_libs, 
                family='binomial')

Psihat.TMLE = TMLE.est$estimates$ATE$`psi`

```

*Answer* The average treatment effect i.e. the counterfactual probability of mortality if all participants recieved blood pressure medication compared to the probability of mortality if all participants did not receive bood pressure medication is 0.326 lower. 

## Discussion Questions

\begin{enumerate}
    \item What is a "double robust" estimator? Why does it provide a guarantee of consistency if either the outcome model or propensity score model is correctly specified? Or in other words, why does mispecifying one of the models not break the analysis? \textbf{Hint}: When answering this question, think about how your introductory statistics courses emphasized using theory to determine the correct outcome model, and in this course how we explored the benefits of matching.
\end{enumerate}

*Answer* A double dobust estimator such as TMLE, produces an unbiased estimate as long as either the outcome or exosure models are correctly specified. This is because TMLE incorporates an additional targeting step that optimizes bias-variance trade off for the target causal paramter (marginal outcome), unlike other estimating mechanisms that optimize bias-variance trade-off for the conditional outcome. 


# LTMLE Estimation

Now imagine that everything you measured up until now was in "time period 1". Some people either choose not to or otherwise lack access to medication in that time period, but do start taking the medication in time period 2. Imagine we measure covariates like BMI, blood pressure, and cholesterol at that time for everyone in the study (indicated by a "_2" after the covariate name). 

## Causal Diagram

Update your causal diagram to incorporate this new information. \textbf{Note}: If your groups divides up sections and someone is working on LTMLE separately from TMLE then just draw a causal diagram even if it does not match the one you specified above.

\textbf{Hint}: Check out slide 27 from Maya's lecture, or slides 15-17 from Dave's second slide deck in week 8 on matching.

\textbf{Hint}: Keep in mind that any of the variables that end in "\_2" are likely affected by both the previous covariates and the first treatment when drawing your DAG.

```{r}
# DAG for TMLE
# DAG for TMLE
dag_tmle<-dagitty('dag {
bb="0,0,1,1"
"BMI T1" [pos="0.628,0.518"]
"BMI T2" [pos="0.755, 0.614"]
"BP Medication T1" [exposure,pos="0.162,0.730"]
"BP Medication T2" [exposure,pos="0.245, 0.830"]
"Blood Pressure T1" [pos="0.413,0.588"]
"Blood Pressure T2" [pos="0.421, 0.714"]
"Cholesterol T1" [pos="0.764,0.429"]
"Cholesterol T2" [pos = "0.8, 0.5"]
"Race/Ethnicity" [pos="0.219,0.224"]
"Sex at birth" [pos="0.669,0.205"]
Age [pos="0.118,0.337"]
Education [pos="0.462,0.288"]
Income [pos="0.167,0.505"]
Mortality [outcome,pos="0.674,0.737"]
"BMI T1" -> "Blood Pressure T1"
"BMI T1" -> "Cholesterol T1"
"BMI T1" -> Mortality
"BP Medication T1" -> Mortality
"Blood Pressure T1" -> "BP Medication T1"
"Blood Pressure T1" -> Mortality
"Cholesterol T1" -> "Blood Pressure T1"
"Cholesterol T1" -> Mortality
"Race/Ethnicity" -> "BMI T1"
"Race/Ethnicity" -> "BP Medication T1"
"Race/Ethnicity" -> "Blood Pressure T1"
"Race/Ethnicity" -> "Cholesterol T1"
"Race/Ethnicity" -> Education
"Race/Ethnicity" -> Income
"Race/Ethnicity" -> Mortality
"Sex at birth" -> "BMI T1"
"Sex at birth" -> "BP Medication T1"
"Sex at birth" -> "Blood Pressure T1"
"Sex at birth" -> "Cholesterol T1"
"Sex at birth" -> Education
"Sex at birth" -> Income
"Sex at birth" -> Mortality
Age -> "Blood Pressure T1"
Age -> "Cholesterol T1"
Age -> Education
Age -> Income
Age -> Mortality
Education -> "BMI T1"
Education -> "BP Medication T1"
Education -> "Blood Pressure T1"
Education -> "Cholesterol T1"
Education -> Income
Education -> Mortality
Income -> "BMI T1"
Income -> "BP Medication T1"
Income -> "Blood Pressure T1"
Income -> "Cholesterol T1"
Income -> Mortality


"BMI T1" -> "BMI T2"
"BMI T1" -> "Cholesterol T2"
"BMI T1" -> "Blood Pressure T2"
"Cholesterol T1" -> "Cholesterol T2"
"Cholesterol T1" -> "Blood Pressure T2"
"Cholesterol T1" -> "BMI T2"
"Blood Pressure T1" -> "Blood Pressure T2"
"Blood Pressure T1" -> "Cholesterol T2"
"Blood Pressure T1" -> "BMI T2"

Age -> "Blood Pressure T2"
Age -> "Cholesterol T2"
Income -> "BMI T2"
Income -> "Blood Pressure T2"
Income -> "Cholesterol T2"
Education -> "BMI T2"
Education -> "Blood Pressure T2"
Education -> "Cholesterol T2"
"Race/Ethnicity" -> "BMI T2"
"Race/Ethnicity" -> "Blood Pressure T2"
"Race/Ethnicity" -> "Cholesterol T2"
"Sex at birth" -> "BMI T2"
"Sex at birth" -> "Blood Pressure T2"
"Sex at birth" -> "Cholesterol T2"

"BP Medication T1" -> "BMI T2"
"BP Medication T1" -> "Cholesterol T2"
"BP Medication T1" -> "Blood Pressure T2"

"BMI T2" -> "BP Medication T2"
"Cholesterol T2" -> "BP Medication T2"
"Blood Pressure T2" -> "BP Medication T2"

"BP Medication T1" -> "BP Medication T2"

"BP Medication T2" -> Mortality

"BMI T2" -> Mortality
"Cholesterol T2" -> Mortality
"Blood Pressure T2" -> Mortality
}')
plot(dag_tmle)
```

## LTMLE Estimation

Use the `ltmle` package for this section. First fit a "naive model" that \textbf{does not} control for the time-dependent confounding. Then run a LTMLE model that does control for any time dependent confounding. Follow the same steps as in the TMLE section. Do you see a difference between the two estimates?


```{r}
## Naive Model (no time-dependent confounding) estimate
A1 <- heart_disease$blood_pressure_medication
A2 <- heart_disease$blood_pressure_medication_2
Y <- heart_disease$mortality
W_naive <- heart_disease %>% select(sex_at_birth,simplified_race,income_thousands,college_educ, blood_pressure, blood_pressure_2, bmi_2, chol_2)
data_naive <- data.frame(W_naive, A1, A2, Y)

ltmle_naive_est <- 
  ltmle(data_naive, 
        Anodes= c("A1", "A2"), 
        Lnodes=NULL, 
        Ynodes="Y", 
        abar=c(1, 1),
        SL.library = sl_libs)

ltmle_naive_est$estimates

## LTMLE estimate
A1 <- heart_disease$blood_pressure_medication
A2 <- heart_disease$blood_pressure_medication_2
W <- heart_disease %>% select(sex_at_birth, simplified_race,income_thousands,college_educ, blood_pressure)
L <- heart_disease %>% select(blood_pressure_2, bmi_2, chol_2)
data_ltmle <- data.frame(W, A1, L, A2, Y)

ltmle_est <- 
  ltmle(data_ltmle, 
        Anodes=c("A1", "A2"), 
        Lnodes=c("blood_pressure_2", "bmi_2", "chol_2"), 
        Ynodes="Y", 
        abar=c(1, 1), 
        SL.library = sl_libs)

ltmle_est$estimates
```
Yes, there is a noticeable difference between the two estimates. The LTMLE estimate when including the time-dependent confounders is 0.246, while the estimate for the naive model without time-dependent confounders is 

## Discussion Questions

\begin{enumerate}
    \item What sorts of time-dependent confounding should we be especially worried about? For instance, would we be concerned about a running variable for age the same way we might be concerned about blood pressure measured at two different times?
\end{enumerate}

*Answer* We should be more concerned about time-dependent confounders that are affected by the previous treatment relative to time-dependent confounders that are not affected by the previous treatment. So, it makes sense to be more concerned about blood pressure measured at two different times since  prior treatment would affect blood pressure but prior treatment would not affect a running variable like age. 

